The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Natural or tidal?

### 2025-09-27 thesis/abstract.tex
a compltet thesis draft

The satellite galaxies of the Milky Way (MW), except the Magellanic
Clouds, are all dwarf spheroidals (dSphs)---gas-free, non-rotating
stellar systems dynamically dominated by dark matter. The light profiles
of dSphs are usually well-approximated by an exponential law, with a
steep outer cutoff. Yet systems like the Sculptor (Scl) and Ursa Minor
(UMi) dSphs have distant member stars, out to \(\sim10\) effective radii
from the centre, indicative of outer deviations from an exponential law.
The origin of these extended profiles is unclear, and may reflect either
the influence of Galactic tides or an innate feature. In this thesis, we
review the existence of extended light profiles in Scl and UMi,
validating the Bayesian membership catalogue of \citet{jensen+2024}. To
evaluate if tides indeed can explain these density features, we conduct
idealized N-body simulation of both galaxies in the tidal field of the
Milky Way. We find that neither dwarf has been subjected to tidal forces
strong enough to create the observed extended light profiles from
initially exponential profiles. One complication is that Sculptor's
orbit is strongly influenced by the presence of the Large Magellanic
Cloud (LMC). Our models still suggest that the combined tides of the LMC
and MW are unable to explain Scl's outer profile. We conclude that the
extended light profiles of Scl and UMi are innate, likely reflecting
their assembly histories. Mergers, accretion events, or episodic star
formation are possible explanations for the observations of extended
light profiles.

>>>>

The satellite galaxies of the Milky Way (MW), except the Magellanic
Clouds, are all dwarf spheroidals (dSphs)---gas-free, non-rotating,
dark-matter dominated stellar systems. The light profiles of dSphs
typically follow an exponential law, with the outer density sharply
declining . Yet systems like the Sculptor (Scl) and Ursa Minor (UMi)
dSphs host member stars out to \(\sim10\) effective radii, indicating
outer deviations from an exponential law. The origin of these extended
profiles is unclear, possibly arising from Galactic tides or intrinsic
properties. In this thesis, we review the evidence for extended light
profiles in Scl and UMi, validating the Bayesian membership catalogue of
\citet{jensen+2024}. To evaluate if tides can produce these density
features, we conduct idealized N-body simulation of both galaxies in the
tidal field of the Milky Way. We find that neither dwarf experiences
tides sufficient to affect their stellar distribution. One complication
is that Sculptor's orbit is strongly influenced by the presence of the
Large Magellanic Cloud (LMC). Our models still suggest that the combined
tides of the LMC and MW are unable to explain Scl's outer profile. We
conclude that the extended light profiles of Scl and UMi not of tidal
origin. They are instead likely innate, possibly explained by past
mergers, accretion events, or episodic star formation.

### 2025-09-27 thesis/data.tex
a compltet thesis draft

For a satellite, the \emph{spatial likelihood} is specified either as a
one- or two-component elliptical exponential profile. A second component
is included only when the preferred amplitude of the second component is
non-zero. The \emph{proper-motion likelihood} quantifies the agreement
of a star's motion with the dwarf galaxy's systemic motion, accounting
for observational uncertainties. The \emph{CMD likelihood} measures the
consistency of a star's \emph{Gaia} \(G\) magnitude and
\(G_{\rm BP}- G_{\rm RP}\) colour with theoretical isochrones for the
galaxy. For the background likelihoods, the spatial likelihood is a
uniform distribution over the field, and the background CMD and PM
likelihoods are constructed empirically from the stars in an annulus far
from the satellite. Each likelihood is normalized as a probability
density over the respective parameter space. The appendix discusses the
details of quality cuts and the likelihood calculation.

@@

To illustrate the extent of Sculptor and Ursa Minor, we highlight
distant, spectroscopically confirmed members with red-outlined stars in
Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}. In particular,
\citet{sestito+2023a}; \citet{sestito+2023b} targeted distant, bright
stars from the J+24 candidate membership list. The most distant stars
confirmed in their work lie at radii \(\gtrsim 10 R_h\) from the centres
of each dwarf. For an exponential profile, 99.95\% of stars fall within
a radius of \(6R_h\), so the mere presence of very distant member stars
provides evidence for deviations from an exponential profile.

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/scl_selection.png}
\caption[Sculptor sample selection]{The distributions of various samples
of \emph{Gaia} stars for Sculptor. We plot light grey points for ``all''
field stars (with consistent parallaxes and reliable photometry and
astrometry), turquoise points for ``CMD+PM'' selected stars
(Eq.~\ref{eq:sel_cmd_pm}), blue squares for the ``fiducial'' sample
(with membership probability \(P_{\rm sat} > 0.2\)), and indigo diamonds
for the ``RV members'' sample. We mark the two far-outlier stars from
\citet{sestito+2023a} with rust-outlined indigo stars. \textbf{Top:}
Tangent plane \(\xi, \eta\). The orange ellipses represents 3 and 6
half-light radii. \textbf{Bottom left:} Colour magnitude diagram in Gaia
\(G\) magnitude versus \(G_{\rm BP} - G_{\rm RP}\) colour. We plot a
Padova \(12\,\)Gyr, \({\rm [Fe/H] }=-1.68\) isochone in orange. The
black bar in the top left represents the median colour error.
\textbf{Bottom right:} Proper motion in declination \(\mu_\delta\) vs RA
\(\mu_{\alpha*}\) (corrected). The orange point marks the systemic
\citet{MV2020b} proper motion. The black cross represents the median
proper motion error.}\label{fig:scl_selection}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/umi_selection.png}
\caption[Ursa Minor sample selection]{Similar to
Fig.~\ref{fig:scl_selection} except for Ursa Minor. We outline
``velocity confirmed'' members outside a radius of \(6R_h\) with red
stars \citep[from][]{sestito+2023b, pace+2020, spencer+2018}. We also
mark the location of Muñoz 1 with a pink
circle.}\label{fig:umi_selection}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/fornax_selection.png}
\caption[Fornax sample selection]{Similar to
Fig.~\ref{fig:scl_selection} except for Fornax. RV measurements are from
\citet{WMO2009}. Fornax does not show the same extended outer halo of
probable members as Sculptor or Ursa Minor despite having many more
stars.}\label{fig:fornax_selection}
\end{figure}

@@

While the density profiles of all samples agree in the inner regions,
they begin to deviate outside \(6 R_h\). Outside this radius, the
fiducial profile extends 1--2 magnitudes below the ``CMD+PM'' selection
background. Because the fiducial profile in this region depends
critically on the spatial likelihood, its shape is vulnerable to the
assumed density profile (an exponential in J+24). We explore this in
more detail in Appendix \ref{sec:density_extra}.

@@

Fig.~\ref{fig:classical_dwarfs_densities} compares the fiducial density
profiles of Sculptor, Ursa Minor, Fornax, and other classical dwarf
galaxies. Of the classicals, we exclude Antlia II, due to the extremely
high background, and Sagittarius, which was not included in J+24. The
density profiles are scaled to match at the half-light radius, taken
from \citet{munoz+2018}. All of the classical dwarfs appear to be well
described by an exponential profile in the inner regions. In the outer
regions, however, Sculptor and Ursa Minor deviate, and show a clear
outer excess over an exponential law (solid black line). These galaxies
are better fit by a Plummer law (dashed black line). The deviation from
an exponential grows outwards, and at \(\sim 8 R_h\), may reach 2 orders
of magnitude. The remainder of this thesis will be devoted to assessing
whether the outer excess shown by Scl and UMi are due to Galactic tides.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_fnx_density_methods.pdf}}
\caption[Sculptor density profiles]{The density profile of Sculptor,
Ursa Minor, and Fornax for different selection criteria, ploted as log
surface dencity versus log elliptical radius. The samples are: ``all''
selects any high quality and parallax-consistent star (grey
circles),``all--background'' subtracts a uniform background density from
the ``all'' profile (orange pentagons), ``CMD+PM'' select stars
according to CMD and PM only (teal triangles), and ``fiducial'' also
includes spatial information (blue squares). We mark the half-light
radius with a vertical dashed line and the background density with the
horizontal grey line. For Ursa Minor, we show the expected location of
Muñoz 1 stars as a horizontal bar (ranging from plus/minus 3 half-light
radii).}\label{fig:scl_observed_profiles}
\end{figure}

>>>>

For a satellite, the \emph{spatial likelihood} is specified either as a
one- or two-component elliptical exponential profile. A second component
is included only when the preferred amplitude of the second component is
non-zero. The \emph{proper-motion likelihood} quantifies the agreement
of a star's motion with the dwarf galaxy's systemic motion, accounting
for observational uncertainties. The \emph{CMD likelihood} measures the
consistency of a star's \emph{Gaia} \(G\) magnitude and
\(G_{\rm BP}- G_{\rm RP}\) colour with theoretical isochrones for the
galaxy. For the background likelihoods, the spatial likelihood is a
uniform distribution over the field, and the background CMD and PM
likelihoods are constructed empirically from the stars in an annulus far
from the satellite. Each likelihood is normalized as a probability
density over the respective parameter space. Appendix
\ref{sec:density_extra} discusses the details of quality cuts and the
likelihood calculation.

@@

To illustrate the extent of Sculptor and Ursa Minor, we highlight
distant, spectroscopically confirmed members with red-outlined stars in
Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}. In particular,
\citet{sestito+2023a}; \citet{sestito+2023b} targeted distant, bright
stars from the J+24 candidate membership list. The most distant stars
confirmed in their work lie at radii \(\gtrsim 10 R_h\) from the centres
of each dwarf. For an exponential profile, 0.05\% of stars fall outside
a radius of \(6R_h\) and less than \(1/10^6\) fall outside
\(10R_h\)---the mere presence of very distant member stars provides
evidence for deviations from an exponential profile.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_selection.png}}
\caption[Sculptor sample selection]{The distributions of various samples
of \emph{Gaia} stars for Sculptor. In each panel, we plot light grey
points for the ``all'' sample of stars, turquoise points for ``CMD+PM''
selected stars (Eq.~\ref{eq:sel_cmd_pm}), blue squares for the
``fiducial'' sample (\(P_{\rm sat} > 0.2\)), and indigo diamonds for the
``RV members'' sample. We mark the two far-outlier stars from
\citet{sestito+2023a} with rust-outlined indigo stars. \textbf{Left:}
Tangent plane \(\xi, \eta\). The orange ellipses represents 3 and 6
half-light radii. \textbf{Top right:} Colour magnitude diagram in Gaia
\(G\) magnitude versus \(G_{\rm BP} - G_{\rm RP}\) colour. We plot a
Padova \(12\,\)Gyr, \({\rm [Fe/H] }=-1.68\) isochone in orange. The
black bar in the top left represents the median colour error.
\textbf{Bottom right:} Proper motion in declination \(\mu_\delta\) vs RA
\(\mu_{\alpha*}\) (corrected). The orange point marks the systemic
\citet{MV2020b} proper motion. The black cross represents the median
proper motion error.}\label{fig:scl_selection}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_selection.png}}
\caption[Ursa Minor sample selection]{Similar to
Fig.~\ref{fig:scl_selection} except for Ursa Minor. We outline
``velocity confirmed'' members outside a radius of \(6R_h\) with red
stars \citep[from][]{sestito+2023b, pace+2020, spencer+2018}. We also
mark the location of Muñoz 1 with a pink
circle.}\label{fig:umi_selection}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/fornax_selection.png}}
\caption[Fornax sample selection]{Similar to
Fig.~\ref{fig:scl_selection} except for Fornax. The isocrone is instead
for a 2Gyr stellar population. RV measurements are from \citet{WMO2009}.
Fornax does not show the same extended outer halo of probable members as
Sculptor or Ursa Minor despite having many more
stars.}\label{fig:fornax_selection}
\end{figure}

@@

While the density profiles of all samples agree in the inner regions,
they begin to deviate outside \(6R_h\). Outside this radius, the
fiducial profile extends 1--2 magnitudes below the ``CMD+PM'' selection
background. Because the fiducial profile in this region depends
critically on the spatial likelihood, its shape is vulnerable to the
assumed density profile (an exponential in J+24). We explore this in
more detail in Appendix \ref{sec:density_extra}.

@@

Our density profiles are robust to alternative methodologies and
magnitude cuts. Density profiles based on J+24 candidates may be biased
by their assumed spatial likelihoods. To address this, we consider a
Bayesian model with a non-parametric spatial likelihood. We also explore
selections using absolute cuts, and the effect of structural assumptions
of the density profile (Appendix \ref{sec:extra_density}). These
methodologies are consistent except when the derived density drops below
the background of satellite-like stars (from the CMD+PM sample). We also
do not find evidence of magnitude biases in the density profiles,
indicating inhomogeneous magnitude-completeness is small. Our density
profiles furthermore agree with photometric surveys and similar
literature (see Section~\ref{sec:extra_density}). We conclude that
selection criteria do not influence our results.

Fig.~\ref{fig:classical_dwarfs_densities} compares the fiducial density
profiles of Sculptor, Ursa Minor, Fornax, and other classical dwarf
galaxies. Of the classicals, we exclude Antlia II, due to the extremely
high background, and Sagittarius, which was not included in J+24. The
density profiles are scaled to match at the half-light radius, taken
from \citet{munoz+2018}. All of the classical dwarfs appear to be well
described by an exponential profile in the inner regions.\footnote{Comparing
  density profiles of dwarf galaxies is complicated by variations in the
  effective depth between galaxies. However, the deviations between Scl,
  UMi and other galaxies is apparent even when data covers all dwarfs.}
In the outer regions, however, Sculptor and Ursa Minor deviate, and show
a clear outer excess over an exponential law (solid black line). These
galaxies are better fit by a Plummer law (dashed black line). The
deviation from an exponential grows outwards, and at \(\sim 8 R_h\), may
reach 2 orders of magnitude. The remainder of this thesis will be
devoted to assessing whether the outer excess shown by Scl and UMi are
due to Galactic tides.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_fnx_density_methods.pdf}}
\caption[Density profiles for different Gaia samples]{The density
profile of Sculptor, Ursa Minor, and Fornax for different selection
criteria, ploted as log surface dencity versus log elliptical radius.
The samples are: ``all'' selects any high quality and
parallax-consistent star (grey circles),``all--background'' subtracts a
uniform background density from the ``all'' profile (orange pentagons),
``CMD+PM'' select stars according to CMD and PM only (turquoise
triangles), and ``fiducial'' also includes spatial information (blue
squares). We mark the half-light radius with a vertical dashed line and
the background density with the horizontal grey line. For Ursa Minor, we
show the expected location of Muñoz 1 stars as a horizontal bar (ranging
from \(\pm3\) times the cluster's half-light
radii).}\label{fig:scl_observed_profiles}
\end{figure}
